---
title: "The grass ecosystem"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The grass ecosystem}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

> **Note (v0.2.0 update):** This vignette describes a workflow from the
> earlier framework version. The current canonical workflow is the
> Report Card via `grass_report(ratings = Y)`. See
> `vignette("reporting-card", package = "grass")`. The content here
> may be retired in a future version.

## One package, one entry point

`grass` is a single package with a single user-facing reporting call:

```{r eval=FALSE}
grass_report(data, spec = grass_spec_binary(), ...)
```

The `spec` argument selects the metric family. Today only
`grass_spec_binary()` returns a usable spec; the other constructors
(`grass_spec_ordinal()`, `grass_spec_multirater()`,
`grass_spec_continuous()`) are placeholders that produce valid spec
objects but error with a pointer to `?grass_roadmap` when passed to
`grass_report()`. This is deliberate: users can read the planned API
and write code that will be ready for the release, without the package
pretending to compute numbers it cannot yet compute.

## Why one package, many families

Inter-rater and test-retest reliability are not one problem. A binary
diagnosis, an ordinal severity scale, a continuous assay, and the
internal consistency of a survey each live in a different statistical
literature with a different reference standard. The `grass` framework
treats them as branches of a common problem — "what value should this
agreement coefficient take given the study's actual parameters?" — and
attaches a prevalence-conditioned reference curve to each.

The alternative, a package per family, fragments the user's workflow
without improving the science. A one-package design with internal
dispatch means:

* Existing user code does not change when a new family lands.
* The same verbs — `grass_methods()`, `grass_format_report()`,
  `plot()`, `tidy()` — apply across every family.
* The user does not have to remember which package exports what.

The framework labels *GRASS*, *TURF*, *MEADOW*, *FIELD* exist in the
paper and in `?grass_roadmap` as organising concepts. They never
appear as function prefixes the user has to learn.

## Today: the binary family

```{r}
library(grass)
tab <- matrix(c(88, 10, 14, 88), nrow = 2,
              dimnames = list(R1 = c("0","1"), R2 = c("0","1")))
r <- grass_report(tab, format = "matrix",
                  spec = grass_spec_binary(reference_level = 0.85))
r
```

`reference_level` is the Se = Sp operating point on the analytical
diagonal. Four bands ship — 0.70, 0.80, 0.85, 0.90 — each defined in
closed form under conditional independence of the two raters given the
true class.

## Manuscript-ready methods paragraph

```{r}
cat(grass_methods(r))
```

The paragraph is GRRAS-compliant (Guidelines for Reporting Reliability
and Agreement Studies), fills in the study's numbers, and is safe to
paste into an `.Rmd` or LaTeX manuscript. When the ordinal, multirater,
or continuous families ship, each implements its own
`grass_methods()` method with the paragraph appropriate to that
design; the user call does not change.

## Tomorrow: the rest of the framework

`?grass_roadmap` lays out the planned families. The release cadence is
one paper and one minor release per family. Every family shares the
core commitment: the interpretation of an agreement coefficient must be
conditioned on the study's actual parameters (prevalence, design,
scale), not applied from a fixed 1977 cutoff table.
