---
title: "Introduction to grass: prevalence-aware agreement interpretation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to grass: prevalence-aware agreement interpretation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 110
)
library(grass)
```

> **Note (v0.2.0 update):** This vignette describes a workflow from the
> earlier framework version. The current canonical workflow is the
> Report Card via `grass_report(ratings = Y)`. See
> `vignette("reporting-card", package = "grass")`. The content here
> may be retired in a future version.

## What `grass` does

`grass` computes Cohen's κ, PABAK, and Gwet's AC1 for binary inter-rater
agreement and reports them alongside their **context** — the prevalence
of the positive class, the bias between raters, and simulation-derived
reference curves that show where each metric tends to land for competent
raters across prevalence.

The explicit design goal is to move practice away from fixed qualitative
labels (Landis-Koch and its siblings) and toward a reporting habit that
includes prevalence and bias alongside the agreement metric.

## A balanced example: Cohen (1960) Table 1

Two raters each classify 200 subjects as 0 or 1 on a single binary trait.

```{r}
tab <- matrix(c(88, 10, 14, 88), nrow = 2,
              dimnames = list(R1 = c("0", "1"), R2 = c("0", "1")))
tab

result <- grass_report(tab, format = "matrix")
result
```

All three metrics agree (all ≈ 0.76) because prevalence is near 0.5 and
bias is negligible. The regime label is `balanced`. Nothing surprising
happens when the 2×2 is symmetric; the three metrics track each other.

## A low-prevalence worked example

A rare positive class. 50 subjects, 5 of them positive.

```{r}
tab_rare <- matrix(c(40, 5, 5, 0), nrow = 2,
                   dimnames = list(R1 = c("0", "1"), R2 = c("0", "1")))
result_rare <- grass_report(tab_rare, format = "matrix")
result_rare
```

Observed agreement is 80%. κ is *negative*, because expected agreement
under marginal independence is higher than what was observed. PABAK reports
0.60. AC1 reports 0.76. Three metrics, three different numbers, same data.

The package reports regime = `prevalence-dominated` and the signed
distance of each metric from the reference curve. The framework asks for
*reporting the three metrics together with PI, BI, and the regime*.
Reporting κ alone would classify these raters as poor; reporting PABAK
alone would classify them as substantial. Neither claim is defensible
on its own.

## The landing plot

```{r, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot(result_rare)
```

Three curves show where κ, PABAK, and AC1 tend to fall across prevalence
for raters at the calibrated quality level. The three points show where
*this* study lands. Reading the plot is qualitative and visual — the user
does the interpretation, the package does the bookkeeping.

## The regime plot

```{r, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot(result_rare, type = "regime")
```

Axes are `PI²` and `BI²`. Points above the diagonal have bias dominating;
points below have prevalence dominating; points near the origin are balanced.
The Feinstein-Cicchetti example sits deep in the prevalence-dominated half.

## Input formats

All four shapes give identical metrics:

```{r}
r1 <- c(rep(0, 40), rep(0, 5), rep(1, 5))
r2 <- c(rep(0, 40), rep(1, 5), rep(0, 5))

grass_compute(list(r1, r2), format = "paired")$values["kappa"]
grass_compute(data.frame(r1 = r1, r2 = r2), format = "wide")$values["kappa"]

long_df <- data.frame(
  subject = rep(seq_along(r1), 2),
  rater   = rep(c("r1", "r2"), each = length(r1)),
  rating  = c(r1, r2)
)
grass_compute(long_df, format = "long")$values["kappa"]
```

Factors and character vectors are coerced to 0/1 silently with a one-time
message identifying which level was mapped to the positive class. Override
with `positive =`.

## What to report in a paper

For any binary inter-rater agreement study:

1. Cohen's κ with its CI (Wilson-logit preferred — it respects the
   `[-1, 1]` bound at small N).
2. PABAK.
3. Gwet's AC1.
4. Prevalence index (PI) and bias index (BI).
5. The regime label.

`as.data.frame(grass_report(...))` gives these as a one-row data.frame
ready to bind to a results table.

## References

- Byrt T, Bishop J, Carlin JB. *Bias, prevalence and kappa.* J Clin Epidemiol. 1993; 46(5):423–429.
- Cohen J. *A coefficient of agreement for nominal scales.* Educ Psychol Meas. 1960; 20(1):37–46.
- Feinstein AR, Cicchetti DV. *High agreement but low kappa: I. The problems of two paradoxes.* J Clin Epidemiol. 1990; 43(6):543–549.
- Gwet KL. *Computing inter-rater reliability and its variance in the presence of high agreement.* Br J Math Stat Psychol. 2008; 61(1):29–48.
- Landis JR, Koch GG. *The measurement of observer agreement for categorical data.* Biometrics. 1977; 33(1):159–174.
