---
title: "GRASS methods companion: notation, calibration detail, and auxiliary diagnostics"
author:
  - Austin Semmel
  - Rachel Gidaro
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GRASS methods companion: notation, calibration detail, and auxiliary diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 110,
  eval = FALSE   # populate to TRUE per-chunk as content lands
)
library(grass)
```

## Scope of this companion

This vignette accompanies the GRASS methods paper. It is a parallel
reference: the paper is self-contained for review, while this companion
carries the procedural detail and auxiliary diagnostics a practitioner
running `grass` may want at hand. Section numbering does not parallel
the paper.

Content moved here from earlier paper appendices during the 2026-05-04
appendix consolidation. **The paper does not cite this vignette**; this
is a free-standing artifact for `grass` users.

---

## 1. Notation glossary

> **Status:** stub. Populate with the full notation table that previously
> opened paper App A.

Symbols used throughout the package and the paper:

- `Y` --- the $N \times k$ binary rating matrix; rows are subjects,
  columns are raters.
- `pi_hat` --- observed marginal positive rate.
- `tau2_hat` --- between-subject variance of the per-subject mean call
  rate (sample-side, computed from `Y`).
- `Se_j`, `Sp_j` --- per-rater sensitivity and specificity, inferred
  from the rating matrix, not against external truth.
- `q` --- diagonal-DGP rater quality; $\mathrm{Se}_j = \mathrm{Sp}_j = q$
  for every rater under the diagonal case the surfaces are calibrated on.
- `q_hat` --- internal scaffolding only. Used by `position_on_surface()`
  to select the sampling distribution from which the percentile is
  read; not exposed on the Report Card.
- `delta_hat` --- cross-coefficient surface-percentile spread.

---

## 2. Recovering $\hat q$ from an observed coefficient

> **Status:** stub. Populate with the procedural detail previously in
> paper App A.6.

For each agreement coefficient, `position_on_surface()` inverts the
closed-form expectation curve to recover $\hat q$ at the observed
coefficient value.

- **PABAK**: algebraic inverse, $\hat q = \tfrac{1}{2}(1 + \sqrt{\mathrm{PABAK}})$.
- **Fleiss $\kappa$, Krippendorff $\alpha$, AC1, ICC**: 501-point
  $q$-grid lookup with linear interpolation between the two nearest
  grid points where the curve crosses the observed value.
- **Out-of-range observations**: clamped to the boundary $q$ value
  with a `clamped = TRUE` flag.

---

## 3. ICC variant selection

> **Status:** stub. Expand the one-paragraph paper App E pointer into a
> practitioner-readable walkthrough.

Variant selection follows \citet{koo2016guideline} on both the
inter-rater and intra-rater axes. The package exposes each ICC variant
through the `metric` argument of `position_on_surface()`. This section
walks through the four-question decision tree (same-rater-set / rater
sampling / type / definition) with a worked example per branch.

---

## 4. Coefficient recovery: the full diagnostic

> **Status:** stub. Populate with figures and tables that didn't make
> the lean App F print version (k=5 vs k=25 best-metric map,
> asymmetry-cost figures across the full grid, oracle-vs-fitted-ICC
> head-to-head detail).

The paper's App F carries the lead recovery table
(`tab:best_metric_fitted_k`) plus the asymmetry-cost figures. This
companion expands the same dataset across the full grid.

---

## 5. $\hat\delta$ operating-characteristic sensitivity

> **Status:** stub. Populate with the sensitivity tables, ROC figures,
> small-$N$ floor caveat, and same-sign comparison detail beyond what
> App G keeps in print.

Paper App G carries the threshold calibration, the adopted-thresholds
table, the by-$k$ summary, and the same-sign vs mixed-sign comparison
table. This companion adds:

- Full sensitivity strip across candidate thresholds (5, 10, 12, 15,
  20 pp) at every $(k, N, A)$ cell.
- Per-$k$ ROC curves at $N = 50, 200, 500$.
- Small-$N$ floor: $k = 2$, $N = 50$ diagonal-null $\hat\delta$
  distribution detail.
- Sign-cancellation detail: same-sign vs mixed-sign vs balanced-sign
  per cell.

---

## 6. Worked code examples per Report Card branch

> **Status:** stub. Populate with three runnable examples that each
> produce the corresponding Report Card output: aligned panel,
> divergent panel, $k = 2$ counterexample.

```{r aligned-example}
# §4.1 aligned case: Se = Sp = 0.85, k = 5, N = 1000, seed = 5
# (TODO: populate with the simulate-and-call sequence and walk through
# the returned grass_card object field by field)
```

```{r divergent-example}
# §4.2 divergent case: op_strong panel at k = 5, N = 1000, seed = 6
# (TODO: populate; show pairwise + pooled-reference + LCA outputs)
```

```{r k2-example}
# §4.3 k=2 counterexample: q = 0.97, k = 2, N = 200, seed = 17
# (TODO: populate; show envelope-ceiling clamp behavior)
```

---

## 7. Audit pass-rate detail

> **Status:** stub. Populate with the per-coefficient audit pass-rate
> tables across (q, F-key, k, N) that previously sat in paper App C.

The paper's App B (Reproducibility, grids, and DGPs) summarizes audit
pass rates at Bonferroni-corrected $|z| < 4.80$. This companion lists
the per-cell audit results for reviewers wanting to verify any
specific corner.

---

## 8. Confidence-interval methods detail

> **Status:** stub. Populate with the procedural detail moved out of
> paper App D — Wilson-logit construction, delta-method SE on $\hat q$
> for reviewer audit, bootstrap CI procedure for per-rater
> $(\hat\Se_j, \hat\Sp_j)$.

The paper App B carries the load-bearing CI numbers (Monte-Carlo error
on percentile, nearest-neighbour cost, bootstrap coverage at one
calibration point). This companion adds the implementation detail.

---

## References

The paper's bibliography is the authoritative source. Key
methodological references this companion leans on:

- Byrt, Bishop, Carlin (1993) — PABAK derivation.
- Cohen (1960), Fleiss (1971) — $\kappa$ family.
- Gwet (2008) — AC1.
- Krippendorff (2011, 2019) — $\alpha$.
- Koo & Li (2016) — ICC variant selection.
- Hui & Walter (1980) — $k = 2$ identifiability bounds.
- Dawid & Skene (1979) — latent-class fit.
- Neyman & Pearson (1933) — size-controlled testing.

(Full citations in `references.bib` next to the paper source.)
