---
title: "GRASS Report Card: a one-call workflow for binary rater reliability"
author:
  - Austin Semmel
  - Rachel Gidaro
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GRASS Report Card: a one-call workflow for binary rater reliability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 110
)
library(grass)
```

## Abstract

Rater-reliability coefficients drift with prevalence, rater count, and
sample size, so a fixed-cutoff label like "substantial agreement" can
land on the same panel and mean different things. This package replaces
fixed cutoffs with a context-conditioned **Report Card**, computed in one
call from the rating matrix alone. The workflow returns four fields: the
sample summary, the primary coefficient with its surface percentile, the
cross-coefficient asymmetry diagnostic, and (when the panel disagrees
with itself) a per-rater latent-class fit.

### What's new in 0.2.2

- Two new transparency rules in the `delta_hat` calculation: clamped
  coefficients (whose observed value falls outside the surface envelope)
  are excluded from `delta_hat` rather than inflating it, and the
  exclusion is surfaced both in the printed Report Card (under the
  `divergent` flag, clamped rows show `[clamped — excluded from delta]`)
  and in the `panel` data.frame's new `clamped` column. See Section 7.
- A new top-level `plot_surface()` function for prospective study design:
  given `metric`, optional `pi_hat`, and optional `observed`, returns
  a heatmap of the closed-form `E[metric](M_1, q)` reference surface
  with optional pin marker. See Section 8.
- When called with `axis = "intra"`, `grass_report()` now surfaces a
  note that the intra-axis surface percentile uses the bundled
  inter-rater diagonal calibration as an approximation; the dedicated
  intra-axis calibration cube is queued for a follow-on data release.

## 1. The headline call

The entire workflow is three lines: load the package, pass an `N x k`
binary rating matrix, print the result.

```{r}
library(grass)
set.seed(2026)

# Tiny demo: 5 raters, 200 subjects, modest agreement.
truth <- rbinom(200, 1, 0.30)
Y <- sapply(1:5, function(j) {
  ifelse(truth == 1, rbinom(200, 1, 0.85), rbinom(200, 1, 0.15))
})

grass_report(ratings = Y)
```

Each block in the printed Report Card means something specific:

- **`sample`** describes the study: rater count `k`, subject count `N`,
  and the observed positive rate `pi_hat = mean(Y)`.
- **The primary coefficient line** (e.g. `PABAK = 0.41 -> 95th percentile`)
  shows the observed value and where it lands on the reference surface
  *for this study's `(k, N, pi_hat)`*. The percentile is calibrated, not
  comparative against fixed cutoffs.
- **`band`** translates the percentile into a plain-language label
  (Poor / Moderate / Strong / Excellent) with a confidence qualifier
  (decisive / moderate / weak).
- **`delta`** is the cross-coefficient percentile spread in pp. Below
  9.25 the panel is *aligned* and one coefficient is safe to cite. Above
  11.75 the panel is *divergent* and the headline coefficient is
  suppressed in favor of per-rater detail.

## 2. A symmetric panel (the easy case)

Five raters with the same accuracy (Se = Sp = 0.85) on a moderate-prevalence
binary task. The panel should land in the *aligned* tier: every coefficient
agrees about where the panel sits, so a single number is enough.

```{r}
set.seed(29)
N <- 200; k <- 5
truth <- rbinom(N, 1, 0.30)
Y_sym <- sapply(1:k, function(j) {
  ifelse(truth == 1, rbinom(N, 1, 0.85), rbinom(N, 1, 0.15))
})

card_sym <- grass_report(ratings = Y_sym)
card_sym
```

Note three things:

1. The `flag` is `aligned` (`delta = 5.8 pp`, comfortably below the
   9.25-pp caution threshold). One coefficient is safe to cite.
2. Only the primary coefficient prints (PABAK, picked automatically by
   the `(k, pi_hat)` rule from Table 2 of the paper). The full panel of
   five coefficients is still computed and stored on the object — it is
   just not the headline.
3. The band (`Strong, decisive`) is a calibrated label, not a Landis-Koch
   inheritance. `Strong` means the surface percentile sits in the third
   quartile of the reference distribution; `decisive` means the modal-band
   sampling probability is concentrated enough to give that label without
   hedging.

## 3. A divergent panel (the killer case)

Now the §4.2 paper case (the `op_strong` profile): five raters with
*alternating* per-rater bias direction at balanced prevalence. Three
raters lean toward over-calling (Se = 0.95, Sp = 0.75); two lean toward
under-calling (Se = 0.75, Sp = 0.95). The asymmetry is severe per
rater but cancels in the panel-aggregate marginals --- so each
agreement coefficient returns roughly the same observed value, yet
their inversions onto the (symmetric-DGP) reference surfaces split
across percentiles. This is the configuration where the headline
`grass_report` flag flips from a single coherent reading to a
per-rater diagnostic.

```{r}
set.seed(6)
N <- 1000; k <- 5
# Alternating per-rater Se/Sp asymmetry: R1, R3, R5 favor sensitivity;
# R2, R4 favor specificity.
Se <- c(0.95, 0.75, 0.95, 0.75, 0.95)
Sp <- c(0.75, 0.95, 0.75, 0.95, 0.75)
truth <- rbinom(N, 1, 0.50)
Y_div <- sapply(1:k, function(j) {
  ifelse(truth == 1, rbinom(N, 1, Se[j]), rbinom(N, 1, 1 - Sp[j]))
})

card_div <- grass_report(ratings = Y_div)
card_div
```

The flag is `divergent`: the cross-coefficient percentile spread
exceeds the 11.75-pp threshold even though the four agreement
coefficients (PABAK, AC1, Fleiss kappa, Krippendorff alpha) all sit
near the same observed value. Three things change in the printed
output:

1. The full panel of percentiles prints, not just the primary, so the
   reader can see that AC1 separates from the kappa-family
   coefficients.
2. The `band` is `suppressed`. Picking a single number from a panel
   whose coefficients disagree on the surface position would hide the
   asymmetry.
3. The per-rater table appears. A latent-class fit (Dawid-Skene EM at
   k >= 3, Hui-Walter bounds at k = 2) recovers each rater's
   sensitivity and specificity with bootstrap confidence intervals.
   The recovered estimates pin the panel's structure: R1, R3, R5 land
   near (Se, Sp) = (0.95, 0.75), R2, R4 near (0.75, 0.95). The
   bootstrap CIs are wide because the latent-class likelihood becomes
   ridge-flat near balanced prevalence and the bootstrap resamples
   both the correct mode and its label-switched counterpart; the
   point estimates pin the local mode, but without external
   orientation the practitioner cannot assert from the rating matrix
   alone which raters are over- versus under-callers.

### Pairwise reliability under the divergent flag

When the panel is divergent the framework's recommended primary
deliverable is `pairwise_agreement()`: a `k x k` matrix of pairwise
PABAK values, each placed on the `k = 2` reference surface at the
pair's observed marginal, plus per-rater behavior against the
panel-majority of the *other* `k - 1` raters as a proxy reference.
The pairwise matrix exposes the panel's structure directly, and the
pooled-reference per-rater table avoids the latent-class
identifiability problem (label-switching is moot because the
"reference" is the observable panel majority, not a latent class).

```{r}
pw <- pairwise_agreement(Y_div)
pw
```

Read the per-rater table: under the §4.2 paper case (R1, R3, R5
favor sensitivity; R2, R4 favor specificity) the pooled-reference
estimates pin each rater's bias direction without ambiguity. R1's
`Se_tilde` is high (calls true positives at panel-majority rate)
while `Sp_tilde` is lower (calls panel-majority negatives less
reliably) --- the over-caller signature. R2 shows the mirror: low
`Se_tilde`, high `Sp_tilde`. The latent-class fit's wide bootstrap
CIs in §4.2 of the paper had to live with the label-switched mode;
the pooled-reference per-rater table sidesteps that ambiguity.

The pairwise PABAK matrix lower triangle shows the actual pairwise
agreements; the upper triangle shows surface percentiles at each
pair's observed marginal. Note that the same PABAK value can land
at different surface percentiles across pairs because each pair
sees a different `pi_hat_ij` --- the divergent flag is exposing
exactly this prevalence-by-pair structure that a panel-aggregate
summary would have averaged away.

## 4. Layered access

The Report Card is the curated headline. Every field underneath the
printed output is available on demand.

```{r}
summary(card_sym)
```

`summary()` is the *audit view*: full panel of coefficients with
percentiles and bands, all caveats from the surface lookup, the GRASS
version, the timestamp. Use it when you want to see what the headline
hides.

```{r}
head(as.data.frame(card_sym), 5)
```

`as.data.frame()` is the *tidy view*: one row per coefficient with
`observed_value`, `surface_percentile`, `band`, `qualifier`, and
`is_primary`. Use it when you want to bind many studies' Report Cards
into one table for a meta-analysis or to feed the panel into another
analysis pipeline.

```{r plot-surface, fig.alt = "Surface position plot"}
plot(card_sym)                       # default: surface with observation pinned
```

```{r plot-thermometer, fig.alt = "Thermometer view"}
plot(card_sym, type = "thermometer") # delta_hat thermometer
```

The default `plot()` view is the *visceral artifact*: a heatmap of the
expected primary coefficient over `(M_1, q)` at the study's `(k, N)`,
with a single point pinned at the observed coordinates. Use it when you
want to show a reader where the study lands. The `thermometer` view
isolates `delta_hat` against the caution and divergent thresholds; useful
when the question is *"is the panel stable enough to cite a single
number?"*. Additional views (`type = "panel"` for the cross-coefficient
spread, `type = "intervals"` for forest-style CIs, `type = "per_rater"`
for the divergent-branch per-rater forest, `type = "diagnostic"` for a
combined patchwork) are documented in `?plot.grass_card`.

## 5. Granular building blocks

For users who want lower-level control, three of the internals
`grass_report()` calls are exported:

```{r}
position_on_surface(ratings = Y_sym, metric = "pabak")
```

`position_on_surface()` does the surface lookup for a single coefficient
and returns the percentile, band, and band-probability vector.

```{r}
check_asymmetry(ratings = Y_sym)
```

`check_asymmetry()` returns the cross-coefficient percentile spread and
the (aligned / caution / divergent) flag without computing per-rater
detail. Cheap when you want only the diagnostic.

```{r}
latent_class_fit(ratings = Y_div, B = 200)
```

`latent_class_fit()` produces the per-rater Se/Sp table directly, with
Dawid-Skene EM and a nonparametric bootstrap CI at k >= 3.

## 6. Two-rater binary case (k = 2)

The headline call works at k = 2 with no signature change:

```{r}
set.seed(42)
N <- 200
truth <- rbinom(N, 1, 0.30)
Y_k2 <- cbind(
  ifelse(truth == 1, rbinom(N, 1, 0.85), rbinom(N, 1, 0.15)),
  ifelse(truth == 1, rbinom(N, 1, 0.85), rbinom(N, 1, 0.15))
)

grass_report(ratings = Y_k2)
```

When the panel goes divergent at k = 2, per-rater Se and Sp are not
point-identified from the rating matrix alone (there is one fewer
constraint than parameter). The latent-class fit returns Hui-Walter
*bounds* rather than point estimates:

```{r}
latent_class_fit(ratings = Y_k2, B = 200)
```

The bounds carry the same meaning as the k >= 3 point estimates: an
honest representation of what the data say about each rater. At k = 2
that is a range; at k >= 3 it is a point with uncertainty.

## 7. Surface-envelope clamp (an edge case worth knowing about)

`delta_hat` measures cross-coefficient surface-percentile spread. If
one of the panel's coefficients reports an observed value that falls
outside the achievable range of its reference surface at the study's
design `(pi_hat, k, N)`, the inversion to `q_hat` clamps to the
surface boundary and the percentile lands at exactly 0 or 100. Since
v0.2.2, `check_asymmetry()` and `grass_report()` *exclude* such
clamped coefficients from `delta_hat` whenever at least two unclamped
coefficients remain — otherwise a single clamped coefficient could
inflate the spread by 100 pp and fire a false `divergent` flag.

The most common trigger is ICC at `N > 200`. The bundled fitted-ICC
reference is calibrated at `N` in `{50, 200}`; beyond that, the
function falls back to the oracle ICC reference (closed-form, no
GLMM-gap correction), which over-predicts the practitioner's
glmer-fit ICC and clamps the percentile to 0:

```{r}
# A panel where ICC clamps but the four agreement coefficients align.
set.seed(5)
N <- 1000L; k <- 5L
mu_logit <- qlogis(0.08)
p_i <- plogis(rnorm(N, mu_logit, sqrt(0.25)))
truth <- rbinom(N, 1, p_i)
Y_clamp <- matrix(0L, N, k)
for (j in seq_len(k)) {
  Y_clamp[, j] <- rbinom(N, 1, ifelse(truth == 1, 0.85, 0.15))
}
card_clamp <- grass_report(Y_clamp, bootstrap_B = 50)
card_clamp
```

The printed Report Card surfaces the clamp in two places: the
`Notes` section names which coefficient was excluded and why, and
under the `divergent` flag the panel display marks each clamped
coefficient row with `[clamped — excluded from delta]`. The `panel`
data.frame on the returned object also carries a `clamped` logical
column for programmatic inspection:

```{r}
card_clamp$panel[, c("coefficient", "surface_percentile",
                     "band", "clamped")]
```

The trade-off: a coefficient with no calibrated reference at the
user's design contributes nothing to the diagnostic. The benefit:
when `delta_hat` does fire `divergent`, it reflects genuine
cross-coefficient surface-percentile disagreement rather than an
envelope artifact. If the clamp is cosmetic (the four agreement
coefficients align cleanly without ICC), the Report Card's
verdict is the right one. If clamping coincides with a real
divergent panel (rare but possible), the `Notes` line and the
`clamped` column let the practitioner audit which coefficients
contributed to the diagnostic.

## 8. Prospective design: `plot_surface()`

The Report Card workflow assumes you already have a rating matrix.
For study design — "what would my reference surface look like at
`k = 8, N = 200, pi_hat = 0.30` before I collect data?" — use the
new `plot_surface()` entry point.

The bare call shows the closed-form expected value of a coefficient
across the entire surface (mean prevalence `M_1` × rater operating
quality `q`). Dotted contours overlay the four-band partition, so
the reader can see which combinations of prevalence and rater
quality land in which band:

```{r plot-surface-bare, fig.width = 7, fig.height = 4.5}
plot_surface("fleiss_kappa")
```

Adding a design context overlays a dashed reference at the planned
marginal:

```{r plot-surface-design, fig.width = 7, fig.height = 4.5}
plot_surface("mean_ac1",
             pi_hat = 0.30,
             k      = 8,
             N      = 200)
```

Adding a hypothetical observed value pins a marker at
`(pi_hat, hat q)`, where `hat q` is recovered by closed-form
inversion against the reference curve at `pi_hat`. This answers
"if I observe Fleiss kappa = 0.50 in a panel at this design,
where on the surface would I sit?":

```{r plot-surface-pinned, fig.width = 7, fig.height = 4.5}
plot_surface("fleiss_kappa",
             pi_hat   = 0.30,
             observed = 0.50,
             k        = 8,
             N        = 200)
```

`plot_surface()` supports the four closed-form non-ICC metrics
(PABAK, Fleiss kappa, mean-pairwise AC1, Krippendorff alpha). ICC
surfaces depend on the full F-shape and the GLMM-gap-corrected
fitted reference; for ICC, build a card with `grass_report()` and
use `plot.grass_card(type = "surface")`.

## References

- Semmel, A. & Gidaro, R. (2026). *GRASS: A context-conditioned
  reporting convention for binary rater reliability.* (Working paper.)
  The package is the computational companion to this work; the four-field
  Report Card and the surface-percentile machinery are documented in
  detail there.
- Hui, S. L. & Walter, S. D. (1980). Estimating the error rates of
  diagnostic tests. *Biometrics* 36(1), 167-171. The k = 2 bounds
  returned by `latent_class_fit()`.
- Dawid, A. P. & Skene, A. M. (1979). Maximum likelihood estimation of
  observer error-rates using the EM algorithm. *Applied Statistics*
  28(1), 20-28. The k >= 3 EM fit.
The (9.25, 11.75) thresholds on `delta_hat` are
Neyman-Pearson-motivated size-alpha cutoffs (Neyman & Pearson 1933):
the smallest threshold whose false-positive rate on the
diagonal-symmetric null does not exceed 0.05 / 0.01 (paper §3.2 /
App G). NP-optimal within the test class of threshold rules on
`delta_hat`; not derived from a likelihood-ratio statistic on the
joint distribution of the panel coefficients.
