---
title: "Worked examples of skewed agreement"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Worked examples of skewed agreement}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4.5, dpi = 110)
library(grass)
```

> **Note (v0.2.0 update):** This vignette describes a workflow from the
> earlier framework version. The current canonical workflow is the
> Report Card via `grass_report(ratings = Y)`. See
> `vignette("reporting-card", package = "grass")`. The content here
> may be retired in a future version.

## Example 1: Extreme prevalence, high PABAK, near-zero kappa

Byrt-Bishop-Carlin regime. N = 125, and 120 of 125 raw decisions fall in
the negative class.

```{r}
tab <- matrix(c(118, 2, 5, 0), nrow = 2,
              dimnames = list(R1 = c("0", "1"), R2 = c("0", "1")))
result <- grass_report(tab, format = "matrix")
result
```

PABAK reports 0.89, κ reports nearly zero, AC1 lands near 0.94. The regime
is `prevalence-dominated`. The raters have essentially never been tested on
the minority class, so high PABAK does not imply substantial agreement —
it implies there is nothing to disagree about. Report all three metrics
together with PI, and the reader can form their own judgment.

## Example 2: Low prevalence, high observed agreement, negative kappa

```{r}
tab <- matrix(c(40, 5, 5, 0), nrow = 2,
              dimnames = list(R1 = c("0", "1"), R2 = c("0", "1")))
result <- grass_report(tab, format = "matrix")
result
```

80% observed agreement, κ = −0.11. A fixed-cutoff scheme like Landis-Koch
would classify these raters as "poor" on the strength of κ alone. The
GRASS report shows why κ is structurally suppressed here: PI = 0.80 and
the regime is prevalence-dominated.

## Example 3: Bias between raters

```{r}
tab <- matrix(c(60, 10, 25, 5), nrow = 2,
              dimnames = list(R1 = c("0", "1"), R2 = c("0", "1")))
result <- grass_report(tab, format = "matrix")
result
```

The off-diagonal asymmetry (25 vs 10) drives a non-trivial bias index. In
this regime the appropriate response is a calibration session between
raters, not more training on hard cases.

```{r, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot(result, type = "regime")
```

## Is the drift from N or from structure?

A natural question when you see a study plotted on a reference curve:
the observed points land *off* the lines — is that because the sample is
small, or because something structural is going on?

Short answer: **small N widens the scatter; prevalence drives the
drift.** The two effects are separable, and you can see the separation
directly. We compare the same data-generating process at small N vs
large N, then the same small N at balanced vs skewed prevalence.

```{r}
# Helper: simulate two raters at Se = Sp = q against a latent truth.
draw <- function(n, p, q, seed = 1) {
  set.seed(seed)
  truth <- rbinom(n, 1, p)
  r1 <- ifelse(truth == 1, rbinom(n, 1, q), rbinom(n, 1, 1 - q))
  r2 <- ifelse(truth == 1, rbinom(n, 1, q), rbinom(n, 1, 1 - q))
  data.frame(r1 = r1, r2 = r2)
}
```

### Same skew, different N

Same prevalence (p = 0.10), same rater quality (Se = Sp = 0.85),
different sample sizes. The observed points stay in roughly the same
spot relative to the reference lines — kappa below, PABAK/AC1 above —
and the spot does not chase zero as N grows. N tightens the scatter
but does not explain the drift.

```{r fig.show = "hold", out.width = "50%", fig.height = 4.2, eval = requireNamespace("ggplot2", quietly = TRUE)}
r_small <- grass_report(draw(n =  50, p = 0.10, q = 0.85), reference = "high")
r_large <- grass_report(draw(n = 500, p = 0.10, q = 0.85), reference = "high")
plot(r_small, labels = "legend") + ggplot2::labs(title = "N = 50")
plot(r_large, labels = "legend") + ggplot2::labs(title = "N = 500")
```

The deltas between observed and reference are driven by the *structure*
of the problem at p = 0.10 (kappa getting squeezed by its chance term),
not by the 50-vs-500 sample size.

### Same N, different prevalence

Now hold N = 50 constant but move prevalence from 0.10 (skewed) to 0.50
(balanced). At balanced prevalence the three metrics collapse toward
each other and toward their reference curves. The earlier "kappa low,
PABAK/AC1 high" pattern vanishes — it was a prevalence effect, not a
noise effect.

```{r fig.show = "hold", out.width = "50%", fig.height = 4.2, eval = requireNamespace("ggplot2", quietly = TRUE)}
r_skew <- grass_report(draw(n = 50, p = 0.10, q = 0.85), reference = "high")
r_bal  <- grass_report(draw(n = 50, p = 0.50, q = 0.85), reference = "high")
plot(r_skew, labels = "legend") + ggplot2::labs(title = "Prevalence = 0.10 (skewed)")
plot(r_bal,  labels = "legend") + ggplot2::labs(title = "Prevalence = 0.50 (balanced)")
```

### Takeaway

When the three metrics land in visibly different positions relative to
their reference curves, that is the algebra of the prevalence regime —
not a signal that you need more data. Increasing N will tighten each
point toward its own reference, but it will not line them up with each
other. Report all three with PI, BI, and the regime label, and the
reader can tell which effect is which.

## What to report

Always together:

1. Cohen's κ (with a CI that respects the `[-1, 1]` bound at small N)
2. PABAK
3. Gwet's AC1
4. Prevalence index (PI)
5. Bias index (BI)
6. The regime label

`grass_format_report(result)` produces a one-line version of the same
information suitable for pasting into a manuscript sentence or a results
table cell.
